Identifying and Correcting Programming Language Behavior Misconceptions

### 2023-11-14 proposal.tex
Updates from Overleaf
title
Clean up

%% abbrev. for RQs
\newcommand{\rqI}{What program-behavior misconceptions apply to SMoL and present (even) in students with prior programming backgrounds?}
\newcommand{\rqII}{What would be a tutoring system that aims to correct the misconceptions, draws on cognitive, educational, or psychological concepts, and seems to be effective?}

@@

\title{Dissertation Proposal}
\author{Kuang-Chen Lu}
\date{\today}

@@

\begin{abstract}
  Misconceptions about core linguistic concepts like mutable
  variables, mutable compound data, and their interaction with scope
  and higher-order functions seem to be widespread. But how do we
  detect them, given that experts have blind spots and may not realize
  the myriad ways in which students can misunderstand programs?
  Furthermore, once identified, what can I do to correct them?

@@

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, I have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\begin{lstlisting}[language=Python]
button_list = []

@@

In the rest of this document, I first define SMoL (\Cref{s:smol}),
then present a literature review of misconceptions
(\Cref{s:background-misconceptions}) and tutoring systems
(\Cref{s:background-tutoring-systems}), and then outline my research
plans for the RQs (\Cref{s:plan-rq1,s:plan-rq2}). The remaining
sections detail my current progress.

@@

Known misconceptions are often detected with a kind of instruments
called \emph{concept inventories}
\citep{hestenesForceConceptInventory1992,taylorComputerScienceConcept2014}.
In terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question
has one correct answer and several wrong ones. However, the wrong ones
are chosen with great care. Each one has been validated so that if a
student picks it, I can quite unambiguously determine \emph{what
misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many settings.
For instance, an educator can use them with clickers to get quick
feedback from a class. If several students pick a specific wrong
answer, the educator not only knows they are wrong, but also has a
strong inkling of \emph{precisely what} misconception that group has
and can address it directly.}

@@

\section{Plan for \textbf{RQ1}: Find Misconceptions}%
\label{s:plan-rq1}

@@

To find misconceptions that ``present (even) in students with prior
programming backgrounds'', I have been collecting data from multiple
populations that very likely have prior experience
(\Cref{s:generality-tutor}). I have the following data sources so far:
\begin{description}
  \item[University 1] students in a ``Principles of Programming
    Languages'' class at Brown University. The class has about 70--75
    students per year. It is not required, so students take it by
    choice. Virtually all are computer science majors. Most are in
    their third or fourth year of tertiary education; about 10\% are
    graduate students. All have had at least one semester of
    imperative programming, and most have significantly more
    experience with it. Most have had close to a semester of
    functional programming. The student work described here was
    required, but students were graded on effort, not correctness.
  \item[University 2] a primarily public university in the US. It is
    one of the largest Hispanic-serving institutions in the country. As
    such, its demographic is extremely different from those whose data
    were used above. The Tutor was used in one course in Spring 2023,
    taken by 12 students. The course is a third-year, programming
    language course. The students are required to have taken two
    introductory programming courses (C++ focused).
  \item[Textbook] an instrument was published on the website of a
    programming languages textbook. Over the course of 8 months,
    several hundreds of people submitted to the instrument. To protect
    privacy, I intentionally do not record demographic information,
    but I conjecture that the population is largely self-learners (who
    are known to use the accompanying book), including some
    professional programmers.
\end{description}

@@

\section{Plan for \textbf{RQ2}: Correct Misconceptions}%
\label{s:plan-rq2}

@@

\section{Preliminary Results}

@@

The current scope tutorial uses the following prompt:
\begin{quote}
  When I see a variable reference, how do I find the corresponding
  declaration, if any?
\end{quote}
This prompt has a few problems:
\begin{itemize}
  \item It is unclear to adapt it to other tutorials.
  \item Students tend to give brief responses possibly because the
  questions are asking for a lot and because students are tired when
  they reach the end of a tutorial.
\end{itemize}

So I plan to change the question format into a notice-and-wonder style
of questions, which is known to be less demanding
(\Cref{s:pedagogic-techniques}). So I plan to replace the
\textbf{scope} prompt with
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|defvar|s work?
    \item How do \lstinline|deffun|s work?
    \item How do variable referencess (e.g., \lstinline|x| in \lstinline|(+ x 1)|) work?
    \item How do \lstinline|defvar|s, \lstinline|deffun|s, and variable references interact with each other?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{response with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{respond with} what you are uncertain about these language constructs.
\end{quote}
I plan to use a similar prompt in later tutorials. For example, in \textbf{mut-vars},
I would want the Tutor to ask
\begin{quote}
  \emph{Think about} the following questions:
  \begin{itemize}
    \item How do \lstinline|set!|s work?
    \item How do \lstinline|set!|s interact with other language constructs (namely, \lstinline|defvar|s, \lstinline|deffun|s, and variable references)?
  \end{itemize}
  If you believe you know the answers perfectly, please \emph{respond with}
  \begin{quote}
    I know the answers with 100\% certainty.
  \end{quote}
  Otherwise, please \emph{respond with} what you are uncertain about these language constructs.
\end{quote}

>>>>

%% abbrev. for RQs
\newcommand{\rqI}{What program-behavior misconceptions apply to SMoL and present even in students with prior programming backgrounds?}
\newcommand{\rqII}{How effective is a tutoring system driven by cognitive and educational principles at correcting these misconceptions?}

@@

\title{Identifying and Correcting Programming Language Behavior Misconceptions}
\author{Kuang-Chen Lu}
\date{\today}

@@

\begin{abstract}
  Misconceptions about core linguistic concepts like mutable
  variables, mutable compound data, and their interaction with scope
  and higher-order functions seem to be widespread. But how do we
  detect them, given that experts have blind spots and may not realize
  the myriad ways in which students can misunderstand programs?
  Furthermore, once identified, what can we do to correct them?

@@

Unfortunately, this combination of features appears to also be non-trivial for
programmers to understand. Consider the following scenario: to create a
calculator, we have to construct a callback that can be attached to
each button. The following Python program seems to achieve the
construction of these callbacks and the act of pushing the buttons in
order:
\begin{lstlisting}[language=Python]
button_list = []

@@

In the rest of this document, I first define SMoL (\Cref{s:smol}),
then present a literature review of misconceptions
(\Cref{s:background-misconceptions}) and tutoring systems
(\Cref{s:background-tutoring-systems}), and then outline my research
plans for the RQs (\Cref{s:plan-rq1,s:plan-rq2}). After that, I describe preliminary results (\Cref{s:preliminary-results} and propose my next steps (\Cref{s:future-work}).

@@

Known misconceptions are often detected with a kind of instruments
called \emph{concept inventories}
\citep{hestenesForceConceptInventory1992,taylorComputerScienceConcept2014}.
In terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question
has one correct answer and several wrong ones. However, the wrong ones
are chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many settings.
For instance, an educator can use them with clickers to get quick
feedback from a class. If several students pick a specific wrong
answer, the educator not only knows they are wrong, but also has a
strong inkling of \emph{precisely what} misconception that group has
and can address it directly.}

@@

\section{\textbf{RQ1}: Find Misconceptions}%
\label{s:plan-rq1}

@@

To find misconceptions that ``present (even) in students with prior
programming backgrounds'', I have been collecting data from multiple
populations that very likely have prior experience
(\Cref{s:generality-tutor}). I have the following data sources so far:
\begin{description}
  \item[University 1] includes students in a ``Principles of Programming
    Languages'' class at Brown University. The class has about 70--75
    students per year. It is not required, so students take it by
    choice. Virtually all are computer science majors. Most are in
    their third or fourth year of tertiary education; about 10\% are
    graduate students. All have had at least one semester of
    imperative programming, and most have significantly more
    experience with it. Most have had close to a semester of
    functional programming. The student work described here was
    required, but students were graded on effort, not correctness.
  \item[University 2] includes students at a primarily public university in the US. It is
    one of the largest Hispanic-serving institutions in the country. As
    such, its demographic is extremely different from those whose data
    were used above. The Tutor was used in one course in Spring 2023,
    taken by 12 students. The course is a third-year, programming
    language course. The students are required to have taken two
    introductory programming courses (C++ focused).
  \item[Textbook] includes visitors to the website of a
    programming languages textbook. Over the course of 8 months,
    several hundreds of people submitted to the instrument. To protect
    privacy, I intentionally do not record demographic information,
    but I conjecture that the population is largely self-learners (who
    are known to use the accompanying book), including some
    professional programmers.
\end{description}

@@

\section{\textbf{RQ2}: Correct Misconceptions}%
\label{s:plan-rq2}

@@

\section{Preliminary Results}%
\label{s:preliminary-results}

@@

The current scope tutorial uses the following prompt:
\begin{quote}
  When I see a variable reference, how do I find the corresponding
  declaration, if any?
\end{quote}
It is unclear to adapt it to other tutorials.

So I plan to change the question format into a more consistent style. So I plan to replace the
\textbf{scope} prompt with
\begin{quote}
  How do variable referencess (e.g., the \lstinline|x| in \lstinline|(+ x 1)|) , \lstinline|defvar|s, and \lstinline|deffun|s interact with each other?
\end{quote}
I plan to use a similar prompt in later tutorials. For example, in \textbf{mut-vars},
I would want the Tutor to ask
\begin{quote}
  How do \lstinline|set!|s interact with other language constructs (namely, \lstinline|defvar|s, \lstinline|deffun|s, and variable references)?
\end{quote}
